---
title: "Time Series Homework 4"
author: "Team 6 - Oliver Chen, Hannah Enck, Jamil Gracia, Ava Klissouras"
date: "2025-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load libraries
```{r}
library(tsibble)
library(tidyverse)
library(ggplot2)
library(forecast)
library(fable)
library(lubridate)
library(fpp3)
library(tseries)
library(fabletools)
```


Read in data
```{r}
energy <- read.csv("https://raw.githubusercontent.com/sjsimmo2/TimeSeries/refs/heads/master/energy_F2025.csv")
```


Roll up data
```{r}
energy_daily <- energy %>%
  mutate(new_datetime = mdy_hm(datetime_beginning_ept),
         date = as_date(new_datetime)) %>%
  group_by(date) %>%
  summarise(daily_total = sum(mw))

energy_rolled <- energy_daily %>%
  mutate(year_month = format(date, "%Y-%m")) %>%
  group_by(year_month) %>%
  summarise(total_daily_avg = mean(daily_total))
```


Train test split
```{r}
#Train: data from August 2018 - August 2023
train <- energy_rolled %>%
  filter(year_month >= "2018-08" & year_month <= "2023-08")

#Validation: data from September 2023 - August 2024
validation <- energy_rolled %>%
  filter(year_month >= "2023-09" & year_month <= "2024-08")

#Test: data from September 2024 - August 2025
test <- energy_rolled %>%
  filter(year_month >= "2024-09" & year_month <= "2025-08")
```


Exponential Smoothing Model (ESM)
```{r}

```


ESM forecast on validation data
```{r}

```


MAE and MAPE of ESM on validation data
```{r}

```


Graph of actual validation data vs ESM forecasted data 
```{r}

```


Seasonal ARIMA model
```{r}

```


Seasonal ARIMA model forecast on validation data
```{r}

```


MAE and MAPE of seasonal ARIMA model on validation data
```{r}

```


Graph of actual validation data vs seasonal ARIMA forecasted data 
```{r}

```
